Lattice Boltzmann study of mass transfer for complicated geometries

### 2012-01-30 Text/introduction.tex
Did a few sketches and continue cleaning it up.

\newcommand{\beq}{\begin{equation}}
\newcommand{\feq}{\end{equation}}
\newcommand{\beqstar}{\begin{equation*}}
\newcommand{\feqstar}{\end{equation*}}                                     
\newcommand{\beqal}{\begin{equation}\begin{aligned}}
\newcommand{\feqal}{\end{aligned}\end{equation}}
\newcommand{\beqalstar}{\begin{equation*}\begin{aligned}}
\newcommand{\feqalstar}{\end{aligned}\end{equation*}}
\newcommand{\vol}{k_L\,a}
\newcommand{\lbubble}{L_{\mathrm{bubble}}}
\newcommand{\lunit}{L_{\mathrm{unit}}}
\newcommand{\lslug}{L_{\mathrm{slug}}}
\newcommand{\ububble}{U_{\mathrm{bubble}}}
\newcommand{\uliq}{U_{\mathrm{liq}}}
\newcommand{\ugas}{U_{\mathrm{gas}}}
\newcommand{\uoutlet}{U_{\mathrm{outlet}}}
\newcommand{\cbubble}{C_{\mathrm{bubble}}}
\newcommand{\cinlet}{C_{\mathrm{inlet}}}
\newcommand{\coutlet}{C_{\mathrm{outlet}}}
\newcommand{\coverall}{C_{\mathrm{overall}}}
\newcommand{\cstar}{C^{*}}
\newcommand{\cmedium}{C_{\mathrm{medium}}}
\newcommand{\volnondim}{\vol \frac{\lunit}{\ububble+\ugas}}

@@

Three possible scenarios of conducting numerical simulations are possible: 
\begin{enumerate}
\item As far as the hydrodynamic flow is periodic, one can make an assumption of the same boundary
conditions applied to the tracer concentration. Therefore, only one unit cell is simulated with
periodic boundary conditions, see Fig. \ref{fig:benchmark}. As well, no tracer leaves a
domain. The situation resembles a situation for the plug flow, where no tracer can leave a unit
domain. We expect such a situation happening for small Capillary numbers $Ca<0.1$, where the film
is thin and saturates fast with the tracer. Though easier to implement, it rises certain critisism
towards the inlet concentration to be equal to the outlet concentration. In the continuous
formulation there is the concentration difference between inlet and outlet involving the volumetric
mass transfer coefficient:
\beq
\volnondim=\ln\frac{\cstar-\cinlet}{\cstar-\coutlet}.
\feq
The concentration $C(t)$ required for calculation the volumetric mass transfer coefficient, Eq.
(\ref{theor:average:concentration:time}), can be taken as the average concentration in the domain:
\beq
C(t)=\frac{\int_{liquid}{C \mathrm{d}V}}{\int{\mathrm{d}V}}.
\end
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/benchmark_middle.eps}\\
\caption{The two-dimensional benchmarks for the  the mass transfer
coefficient (bottom) for the bubble located near the entrance (top) and at the middle of the domain
(bottom). \label{fig:benchmark}}
\end{figure}
\item The second approach is the same as the first one but the concentration is taken as in
\cite{vanbaten-circular}. The characteristic concentration is taken as the inlet/outlet
concentration:
\beqal
\cinlet=\frac{\int{U(y) C(0,y,t) \mathrm{d}y}}{\int{U(0,y) \mathrm{d}y}}\\
\coutlet=\frac{\int{U(y) C(\Lunit,y,t) \mathrm{d}y}}{\int{U(\Lunit,y) \mathrm{d}y}}\\
\cinlet=\coutlet.
\feqal
Therefore the assumptions  of this approach is that the difference between inlet and outlet is not
considerably large and the tracer is well mixed in the slug.

\item
The approach of \citet{vanbaten-circular} is taken. Periodic boundary conditions were used. By the
definition the mass transfer coefficient is
proportional to the gain of the mass in the system divided by the concentration difference
multiplied on the surface area:
\beq
\vol=\frac{\dot{m}}{P \Delta C}\frac{P}{V}=\frac{\dot{m}}{V (\cstar-C(t))},
\feq
where mass difference in the domain can be calculated as:
\beq
\dot{m}=\frac{m_2-m_1}{t_2-t_1}=\frac{\int_{liq}{C(t_2)\mathrm{d}V}-\int_liq{C(t_1)\mathrm{d}V}}{
t_2-t_1 } .
\feq
In the approach of \citet{vanbaten} the inlet and outlet concentrations were taken as the
characteristic concentration $C(t)$.

\item 
The tracer picking up procedure can be introduced
in the manner of the considerations above. We impose the concentration at the entrance to be
constant $C(0,y)=C_{inlet}$, at the outlet we impose the outflow boundary condition $\frac{\partial
C}{\partial x}(L,y)=0$. To be compliant with the considerations above one needs to impose as well
the outflow boundary conditions for the inlet as well. However, without loss of the generality one
can impose $C_{inlet}=0$, since the walls in the reference frame
moving with the bubble are moving from the inlet to the outlet. That means that no tracer
concentration leaves domain from the inlet side, which is true for the capillary numbers larger
than $0.6$.  However, this boundary condition can not be used in the case of small capillary
numbers, since there is a strong circulation in front of the bubble. One can see the sketch of the
approach in Fig. \ref{fig:benchmark:jos}. The longer the numerical domain length the better will be
results. One can estimate the volumetric mass transfer coefficient in this case through the equation
(\ref{small:conc:mass:estimation}) for small concentrations. Note a few considerations:
\begin{itemize}
\item  The considerations above were established for the homogenous velocity field. In the case of
the non-homogenous velocity field one needs to obtain a concentration from the mass flux:
\beq
C_{outlet}=\frac{\int{U(y) C(L_{unit},y) \mathrm{d}y}}{\int{U(y) \mathrm{d}y}}.
\feq
\item Note, that one cannnot use the definition of the mass transfer coefficient as far as it
varies through the bubble due to inhomogenoities of the flow. Therefore, the volumetric mass
transfer coefficient is introduced that takes into the account the bubble geometry.
\end{itemize}
\begin{figure}
\includegraphics[width=\textwidth]{Figures/benchmark_jos.eps}
\caption{The numerical mass transfer benchmark. \label{fig:benchmark:jos}}
\end{figure}

\end{enumerate}

The numerical benchmark is presented in Fig. \ref{fig:benchmark}. In
the same figure, one can see the proposed benchmark for the tracer. The
physical
picture for the proposed numerical benchmark is  related to classical experiments conducted by
\citet{bercic-mass}. In their work authors studied gas-liquid mass transfer coefficient by
injecting methane as the gas bubble and measuring the dissolved concentration of methane in water
along the channel. This situation implies the periodic boundary conditions and steady-state motion
of bubbles.

\section{Validation}

\section{Lattice Boltzmann Benchmarks}
\citet{vanbaten-circular} represented a bubble with two hemisperes and a thin film. We want to
perform the benchmarking against these two representative cases.

@@

&\frac{\partial C}{\partial \frac{x}{H}} \frac{U_0}{H} \Bigl(\frac{y}{H}\Bigr)^2=\frac{D}{H^2}\frac{\partial^2 C}{\partial \frac{y^2}{H^2}}\\

@@

Still the equation cannot be solved with the separation of variables as it requires the homogenuos boundary conditions. Thus, let us introduce the new variable to be $\Theta=C-C_{wall}$. The system will be changed but not considerably:

@@

This equation can be solved by separation of variables if the solutions is represented as the multiplication of two functions $\Theta(\zeta,\xi)=X(\zeta)Y(\xi)$. The differential equation will become as follows:

@@

>>>>

\newcommand{\beq}{\begin{equation}}
\newcommand{\feq}{\end{equation}}
\newcommand{\beqstar}{\begin{equation*}}
\newcommand{\feqstar}{\end{equation*}}                                     
\newcommand{\beqal}{\begin{equation}\begin{aligned}}
\newcommand{\feqal}{\end{aligned}\end{equation}}
\newcommand{\beqalstar}{\begin{equation*}\begin{aligned}}
\newcommand{\feqalstar}{\end{aligned}\end{equation*}}
\newcommand{\vol}{k_L\,a}
\newcommand{\lbubble}{L_{\mathrm{bubble}}}
\newcommand{\lunit}{L_{\mathrm{unit}}}
\newcommand{\lslug}{L_{\mathrm{slug}}}
\newcommand{\ububble}{U_{\mathrm{bubble}}}
\newcommand{\uliq}{U_{\mathrm{liq}}}
\newcommand{\ugas}{U_{\mathrm{gas}}}
\newcommand{\uoutlet}{U_{\mathrm{outlet}}}
\newcommand{\cbubble}{C_{\mathrm{bubble}}}
\newcommand{\cinlet}{C_{\mathrm{inlet}}}
\newcommand{\coutlet}{C_{\mathrm{outlet}}}
\newcommand{\coverall}{C_{\mathrm{overall}}}
\newcommand{\cstar}{C^{*}}
\newcommand{\cmedium}{C_{\mathrm{medium}}}
\newcommand{\volnondim}{\vol \frac{\lunit}{\ububble+\ugas}}
\newcommand{\omegaplus}{\omega_{+}}
\newcommand{\omegaminus}{\omega_{-}}

@@

Three possible scenarios of conducting numerical simulations are possible: 
\begin{enumerate}
\item As far as the hydrodynamic flow is periodic, one can make an assumption of the same boundary
conditions applied to the tracer concentration. Therefore, only one unit cell is simulated with
periodic boundary conditions, see Fig. \ref{fig:benchmark}. As well, no tracer leaves a
domain. The situation resembles a situation for the plug flow, where no tracer can leave a unit
domain. We expect such a situation happening for small Capillary numbers $Ca<0.1$, where the film
is thin and saturates fast with the tracer. Though easier to implement, it rises certain critisism
towards the inlet concentration to be equal to the outlet concentration. In the continuous
formulation there is the concentration difference between inlet and outlet involving the volumetric
mass transfer coefficient:
\beq
\volnondim=\ln\frac{\cstar-\cinlet}{\cstar-\coutlet}.
\feq
The concentration $C(t)$ required for calculation the volumetric mass transfer coefficient, Eq.
(\ref{theor:average:concentration:time}), can be taken as the average concentration in the domain:
\beq
\label{average:concentration:definition}
C(t)=\frac{\int_{liquid}{C \mathrm{d}V}}{\int{\mathrm{d}V}}.
\feq
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/benchmark_middle.eps}\\
\caption{The two-dimensional benchmarks for the  the mass transfer
coefficient (bottom) for the bubble located near the entrance (top) and at the middle of the domain
(bottom). \label{fig:benchmark}}
\end{figure}
\item The second approach is the same as the first one but the concentration is taken as in
\cite{vanbaten-circular}. The characteristic concentration is taken as the inlet/outlet
concentration:
\beqal
&\cinlet=\frac{\int{U(y) C(0,y,t) \mathrm{d}y}}{\int{U(0,y) \mathrm{d}y}}\\
&\coutlet=\frac{\int{U(y) C(\lunit,y,t) \mathrm{d}y}}{\int{U(\lunit,y) \mathrm{d}y}}\\
&\cinlet=\coutlet.
\feqal
Therefore the assumptions  of this approach is that the difference between inlet and outlet is not
considerably large and the tracer is well mixed in the slug.

\item
The approach of \citet{vanbaten-circular} is taken. Periodic boundary conditions were used. By the
definition the mass transfer coefficient is
proportional to the gain of the mass in the system divided by the concentration difference
multiplied on the surface area:
\beq
\vol=\frac{\dot{m}}{P \Delta C}\frac{P}{V}=\frac{\dot{m}}{V (\cstar-C(t))},
\feq
where mass difference in the domain can be calculated as:
\beq
\dot{m}=\frac{m_2-m_1}{t_2-t_1}=\frac{\int_{liq}{C(t_2)\mathrm{d}V}-\int_liq{C(t_1)\mathrm{d}V}}{
t_2-t_1 } .
\feq
In the approach of \citet{vanbaten-circular} the inlet and outlet concentrations were taken as the
characteristic concentration $C(t)$.

\item 
The approach of the opened boundaries was examined, see Fig. \ref{fig:benchmark:jos}. At the inlet
and outlet boundaries the diffusion flux was imposed to be zero, $\frac{\partial C}{\partial x}=0$.
The characteristic concentration was taken as the average concentration in the whole domain.
\begin{figure}
\includegraphics[width=\textwidth]{Figures/benchmark_jos.eps}
\caption{The numerical mass transfer benchmark with open boundaries. The mass transfer is
calculated according to Eq. \ref{theor:average:concentration:time} with the characteristic
concentration accoring to Eq. \ref{average:concentration:definition}. \label{fig:benchmark:jos}}
\end{figure}

\item 
The last approach to be examined in this paper is the approach of the simulation of a lot of unit
volumes, see Fig. \ref{big:benchmar:alot}. This situation corresponds to the simulation of the
bubble train head, after injecting to the pipe and travelling along the pipe. The argument here is
that behind a certain number of the bubble train head, the influence of the boundaries will be
diminished. Thus, the average concentration in time will be governed by Eq.
\ref{theor:continuous:mass:transfer}. This simulation statement is of most correspondence to
experiments and should give right answer. However, one of the
disadvantages is to simulate a certain number of unit cells, thus increasing a computational
complexity.
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/benchmark_alot.eps}\\
\caption{Benchmark for a lot of units. \label{fig:benchmark:alot}}
\end{figure}
\end{enumerate}

Before all the approaches will be examined we will perform benchmarks of parts of geometries that
form a problem.

\section{Validation}
As for analytical expressions the problem of the mass transfer can be decomposed to the following
parts: the mass transfer from two hemispheres and the mass transfer from the film. We will examine
these problems with the help of the lattice Boltzmann method and compare them against analytical
solutions. The next sections will present the lattice Boltzmann method and a few benchmarks.

\subsection{TRT $D2Q9$ model}
The lattice Boltzmann equation (LBE) operates on a rectangular grid representing the
physical domain. It utilizes
probability distribution functions (also known as particle populations)
containing information about
macroscopic variables, such as fluid density and momentum. LBE consists of
two parts: a local collision step, and a propagation step which transports
information from one node to another along some 
directions specified by the discrete velocity set.
The LBE is typically implemented as follows:
\begin{equation}
\label{standard:implementation}
\begin{aligned}
&f_i^{*}(\bm{x},t)= f_i(\bm{x},t)-\omega \bigl(f_i(\bm{x},t)-f_i^{eq}(\bm{x},t)\bigr),&&\text{
collision step}\\
&f_i(\bm{x}+\bm{c_i},t+1)=f_i^{*}(\bm{x},t),&&\text{ propagation step}, 
\end{aligned}
\end{equation}
where $f_i$ is the probability distribution function in the direction $\bm{c_i}$,
 $f_i^{eq}$ is the equilibrium probability distribution function, $\omega$ is the
relaxation parameter. Collision operator, $-\omega \bigl(f_i- f_i^{eq}\bigr)$ is so-called BGK
collision operator \cite{bgk}. However, the approach to be used here is the TRT
(two-relaxation-times) collision operator \cite{ginzburg-main,ginzburg-saturated-flow}. In
comparison with the widely used BGK collision operator, TRT collision operator have better accuracy
for diffusion and convection fluxes, larger range of parameters where the LBM scheme is stable.

The TRT collision operator \cite{ginzburg-boundary-main}
decomposes the populations and the equilibrium
distribution into the symmetric and antisymmetric parts:
\begin{equation}
\label{trtdecomp}
f^{\pm}_i=\frac{f_i\pm f_{\bar{i}}}{2}\;,\; 
{eq_i}^{\pm}=\frac{eq_i\pm eq_{\bar{i}}}{2}\;,
\end{equation}
where $\bar{i}$ is the opposite direction to the $i$-th direction.
The collision is performed with two independent relaxation rates for 
symmetric and antisymmetric modes:
\begin{equation}
\label{trt}
\begin{aligned}
&f_i^{*}(\bm{x},t)=f_i(\bm{x},t)-\omegaplus (f_i^{+} - eq_i^+)-\omegaminus
(f_i^{-} -
eq_i^-)\\
&f_i(\bm{x}+\bm{c_i},t+1)=f_i^{*}(\bm{x},t).
\end{aligned}
\end{equation}
Note that the BGK collision operator is the particular subclass of the TRT relaxation operator with
$\omegaplus=\omegaminus$. In comparison with the BGK collision operator,
the TRT collision operator has the additional degree of freedom. Thus, the TRT operator then
introduces
so-called magic (free) parameter
$\Lambda=\Bigl(\frac{1}{\omegaplus}-\frac{1}{2}\Bigr)\Bigl(\frac{1}{\omegaminus}-\frac{1}{2}
\Bigr)$. 
This free parameter controls the effective location of the bounce-back
walls \cite{ginzburg-multireflection}, second-order accuracy of
boundary \cite{ginzburg-boundary-main} and interface schemes \cite{ginzburg-discontinious}, 
spatial accuracy \cite{ginzburg-recurrence,servan-trt-stability},
consistency \cite{ginzburg-brinkman} and, to some extent,
stability \cite{kuzmin-stability-optimal,kuzmin-d1q3,servan-trt-stability}.
In particular, $\Lambda=\frac{1}{4}$ achieves the optimal stability for the
linear advection-diffusion isotropic equation \cite{kuzmin-stability-optimal}.

The parameters $\omegaplus$, $\omegaminus$ and $f_i^{eq}$ fully define the lattice Boltzmann
procedure. The two-dimensional nine velocities LBM $D2Q9$ is defined on the set of lattice
velocities with coordinates:
\beqal
&c_{ix}=\{0,1,0,-1,0,1,-1,-1,1\},\text{ for } i=0\dots8\\
&c_{iy}=\{0,0,1,0,-1,1,1,-1,-1\},\text{ for } i=0\dots8\,.
\feqal

The equilibrium functions for $D2Q9$ TRT model are represented as \cite{kuzmin-stability-optimal}:
\begin{equation}
\begin{aligned}
&eq_i^{+}=eq_i^{(m)}+g^{(u)} eq_i^{(u)}\\
&eq_i^{(m)}=t_i^{(m)} c_e+ eq_i^{(a)}\\
&eq_i^{(u)}=t_i^{(u)} \frac{u_x^2+u_y^2}{2}+\frac{u_x^2-u_y^2}{4} p_i^{(xx)}+g_{xy}^{(u)}\frac{u_x
u_y}{4} p_i^{xy}\\
&eq_i^{(a)}=\frac{D_{xx}-D_{yy}}{4} p_i^{xx}+\frac{D_{xy}}{4} p_i^{(xy)},
\end{aligned}
\end{equation}
where the tensor $p_i^{(xx)}=c_{ix}^2-c_{iy}^2$, the tensor  $p_i^{(xy)}=c_{ix} c_{iy}$, the weights
$t_i^{(u,m,a)}$ can be chosen based on stability criteria. Though not the best set of weights for
stability, but the most spread set of weights, so called ``hydrodynamic`` weights, was chosen:
\begin{equation}
t_i^{(u)}=t_i^{(m)}=t_i^{(a)}=\Bigl\{0,\frac{1}{3},\frac{1}{3},\frac{1}{3},\frac{1}{3},\frac{1}{12},
\frac {1}{12},\frac{1}{12},\frac{1}{12}\Bigr\}
\end{equation}

It can be shown through the Chapman-Enskog procedure \cite{chapman}, that the simple update rule
with the equilibrium function presented above restores the anisotropic
advection-diffusion equation:
\beq
\partial_t C+ \partial_{\alpha} C u_{\alpha}=\partial_{\alpha\beta} D_{\alpha\beta} C,
\feq
 with the
following diffusion tensor:
\begin{equation}
D_{\alpha\beta}=
\begin{pmatrix}
D_{xx} + (g^{(u)}-1) u_x^2 & D_{xy}+(g_{xy}^{(u)}-1)u_x u_y\\
D_{xy} + (g_{xy}^{(u)}-1) u_x u_y& D_{yy}+(g^{(u)}-1) u_y^2 
\end{pmatrix}
\end{equation}
To consider the situation when the resolution is not needed to be changed in $x$ and $y$
directions, one can obtain that $D_{xx}=D_{yy}$. The cross-diagonal components of the diffusion
tensor are taken as zeros, i.e. $D_{xy}=0$. In comparison with $D2Q5$ model, with the $D2Q9$ it is
possible to cancel the numerical diffusion by the proper choice
of the equilibrium functions, i.e. $g_{xy}^{(u)}=g^{(u)}=1$. If one introduces

As far as the simulated diffusion tensor is close to $0$ the stability limits restrict the velocity
amplitude to be simulated \cite{kuzmin-stability-optimal}. However, particular choice of weights
can drastically help. Fig. \ref{stability:d2q9} shows the stability limits for D2Q9 with removed
numerical diffusion and with it for two particular choices of weights.
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/d2q9_stability.eps}
\caption{The non-negativity areas and stable sub-domains of the OTRT schemes ($\Lambda=\frac{1}{4}$)
are plotted for the $D2Q9$ (standard weights, left) and the $D2Q9$ (uniform weights, right) when the
numerical diffusion is removed, partially or completely. Two non-negativity areas where all the
equilibrium populations are positive have curvilinear boundaries: they are smaller when
$g^{(u)}_{xy}=0$ (yellow) and than when $g^{(u)}_{xy}=1$ (green+yellow). The non-negativity
condition is required by BGK collision operator. The stable sub-domains for $g^{(u)} = 1,
g^{(u)}_{xy} = 0$ are small (red+green+yellow) triangles. When the numerical diffusion is removed
the stable area is bounded by magenta and the (blue) domain above. The proper choice of weights can
allow to obtain  total triangle $0 \leq u_x^2+u_y^2 \leq 1−c_e$ (the
uniform weights only).
\label{stability:d2q9}}
\end{figure}

The set of equations (\ref{set:equilibrium:binary}) restores the
macroscopic
fluid equations as:
\begin{equation}
\begin{aligned}
&\partial_t \rho+ \partial_{\alpha} \rho u_{\alpha}=0\\
\end{aligned}
\label{advection:diffusion:equation}
\end{equation}
where $D=c_s^2 (\tau-1/2)$ is the viscosity, 
$P_{\alpha\beta}=\Bigl(p_0-k\phi \Delta \phi -\frac{k}{2}|\nabla \phi|^2\Bigr)\delta_{\alpha\beta}
+ k \partial_{\alpha} \phi \partial_{\beta} \phi$  \cite{pooley-contact}.

{\color{red}
While the lattice Boltzmann system has parameters such as the surface tension, the gas and liquid
viscosities, etc., those parameters are not the representative and
proportional quantities of the parameters in a physical system. The parameters of the lattice
Boltzmann are connected with the physical parameters only through the non-dimensional
numbers governing the physics of the problem. In our case, these numbers are the capillary number
and the
viscosity ratio $\frac{\mu_{\mathrm{liq}}}{\mu_{\mathrm{gas}}}$, which are
obtained from the physical world and then matched through the lattice Boltzmann quantities.
Therefore, one can substitute any quantity, i.e.
$U_{\mathrm{bubble}}$, in the physical units or in the lattice Boltzmann units as soon as the
capillary number is the same in both worlds.}

@@

&\frac{\partial C}{\partial \frac{x}{H}} \frac{U_0}{H}
\Bigl(\frac{y}{H}\Bigr)^2=\frac{D}{H^2}\frac{\partial^2 C}{\partial \frac{y^2}{H^2}}\\

@@

Still the equation cannot be solved with the separation of variables as it requires the homogenuos
boundary conditions. Thus, let us introduce the new variable to be $\Theta=C-C_{wall}$. The system
will be changed but not considerably:

@@

This equation can be solved by separation of variables if the solutions is represented as the
multiplication of two functions $\Theta(\zeta,\xi)=X(\zeta)Y(\xi)$. The differential equation will
become as follows:

@@

\section{Lattice Boltzmann Benchmarks}
